\large{\bf Looking forwards and backwards through locally regulated populations }

### 2023-03-09 forwards-and-backwards.tex
up until MMT

\author{ \begin{small}
\begin{tabular}{ll}                              
Alison M. Etheridge 
 & Thomas G. Kurtz \\   
Department of Statistics & Departments of Mathematics and Statistics\\       
Oxford University & University of Wisconsin - Madison \\                   
24-29 St Giles & 480 Lincoln Drive\\                                                         
Oxford OX1 3LB & Madison, WI  53706-1388\\
UK & USA \\                        
etheridg@stats.ox.ac.uk & kurtz@math.wisc.edu     \\
\url{http://www.stats.ox.ac.uk/~etheridg/} & 
\url{http://www.math.wisc.edu/~kurtz/}  \\       \\
\\
Ian Letter&  Peter L. Ralph 
\\   
Department of Statistics & Department of Mathematics \\
Oxford University &University of Oregon\\                   
24-29 St Giles & Fenton Hall\\
Oxford OX1 3LB & Eugene, OR 97403-1222\\
UK & USA \\
restucci@stats.ox.ac.uk  & plr@oregon.edu \\
\url{https://www.stats.ox.ac.uk/~restucci/}&
\url{https://math.uoregon.edu/profile/plr} \\
\\
Terence Tsui 
 &  \\   
Department of Statistics & \\
Oxford University & \\                   
24-29 St Giles& \\
Oxford OX1 3LB & \\
UK & \\
terence.tsui@sjc.ox.ac.uk &      \\
\url{https://www.maths.ox.ac.uk/people/terence.tsui}&  \\
\end{tabular}
\end{small}}

@@

We introduce a broad class of mechanistic spatial models to describe how spatially 
heterogeneous populations live, die, and reproduce. Birth and death rates can depend 
both on spatial position and local population density, defined at a 
spatial position $x$ to be the convolution 
of the point measure encoding the distribution of individuals in the population
with a suitable non-negative integrable kernel
centred on $x$. We pass to three different scaling limits:
an interacting superprocess, a nonlocal partial differential equation (PDE), 
and a classical PDE. 
The classical PDE is obtained both by a two-step convergence argument, in which we 
first scale time and population size and
pass to the nonlocal PDE, 
and then scale the kernel that 
determines local population density; and 
in the important special case in which the limit is a 
reaction diffusion equation, directly 
by simultaneously scaling the kernel and the
timescale and population size in our individual based model.

A novelty of our model is that in describing the dynamics we explicitly
model a juvenile phase: the number of juveniles produced by an individual depends
on local population density at the location of the parent. 
Juvenile offspring are 
thrown off in a (possibly heterogeneous, anisotropic) Gaussian distribution 
around the location of the parent; they then reach (instant) maturity with a probability
that can depend on the local population density (possibly measured through convolution
with a different kernel from that used in determining fecundity) 
at the location at which they land. 
Although we only record mature individuals, a trace of this two-step description 
remains in our population models, resulting in novel limits in which the spatial
motion is governed by a nonlinear diffusion.

@@

Reproduction of individuals naturally leads to spatial branching process models,
including branching random walk, 
branching Brownian motion, and the Dawson-Watanabe superprocesses.  
However, as a result of the branching assumption (once born, individuals behave 
independently of one another), 
a population evolving according to any of these models will
either die out or grow without bound and, in so doing, it can
develop clumps of arbitrarily large density and extent. 
Our starting point here is an individual-based model of a single species 
in continuous space in which
birth, death, and establishment may all depend on local population density
as well as on spatial location,
allowing for stable populations through density-dependent feedback.

Although it is often mathematically convenient to assume that individuals follow
Brownian motion during their lifetime, 
in our model, offspring
are thrown off according to some 
spatial distribution centred on the location of the parent.   
This is particularly appropriate for modelling plant populations, in which
this dispersal of offspring around the parent is the only source of
spatial motion.

Most models do not distinguish between juveniles and adults, so,
for example, the number of adults produced by a single parent is determined
by the degree of crowding at the location of the parent. A novelty
of the models that we introduce here is that, although we shall only
follow the adult population, in formulating the dynamics of the
models we shall distinguish
between production of juveniles, which will depend upon the location of 
the adult, and their successful establishment, which will depend on the
location in which a juvenile lands. The result is that not only the absolute 
number, but also the spatial distribution
around their parent, 
of those offspring that survive to adulthood
will depend upon the local population 
density.

We shall consider three different classes of scaling limits for our model.
The first yields a class of (generalised) superprocesses in which coefficients 
governing both the spatial motion and the branching components of the process can depend
on local population density; the second is a corresponding class of deterministic
non-local differential equations; and the third are classical PDEs.
We measure local population density around a point by convolving with
a smooth kernel $\rho(\cdot)$, which may differ
for the two stages of reproduction. 
When the limit is deterministic,
it is a (weak) solution of an equation of the form
\begin{equation}
	\label{general deterministic limit}
        \partial_t \varphi_t(x)
        =
        r\left(x, \varphi_t \right)
        \DG^* \left[
            \varphi_t(\cdot)
            \gamma\big( x, \varphi_t \big)
        \right](x)
        +
        \varphi_t(x)
        F\left(x, \varphi_t \right)
        ,
\end{equation}
where $\DG^*$ is (the adjoint of) a strictly uniformly elliptic
second order differential operator, typically
the Laplacian.
We shall be more specific about the parameters below.

By replacing $\rho$ by
$\rho^\epsilon(\cdot)=\rho(\cdot/\epsilon)/\epsilon^d$, 
we can also scale the ``width'' of the 
region over which we measure local population density. 
When the population follows~(\ref{general deterministic limit}), 
we can expect that taking a second limit by scaling the kernels appearing
in $r$, $\gamma$, and $F$ in this way, so that
in the limit as $\epsilon\to 0$
interactions are pointwise, we should
recover a nonlinear PDE.
We verify that this is indeed the case in two important examples:
a special case of the porous medium equation with a logistic growth term, in which we 
retain a signal of our two stage approach to modelling in the nonlinear diffusion term
and the limiting equation takes the form 
\begin{equation}
	\label{PME1}
	\partial_t \varphi = \Delta (\varphi^2)+\varphi(1-\varphi);
\end{equation}
and a wide class of semi-linear PDEs of the form 
\begin{equation}
\label{semilinear PDE}
\partial_t \varphi = \DG^*\varphi+ \varphi F(\varphi),
\end{equation}
which includes the Fisher-KPP equation and the Allen-Cahn equation.

It is of interest to understand under what conditions we can replace the two-step limiting
process described above by one in which we simultaneously scale the 
kernels
and the other parameters in 
our population model to arrive at the PDE limit.
This is mathematically much more challenging, but we establish such 
one-step convergence in cases for which the limit is a classical 
reaction diffusion equation of the form~(\ref{semilinear PDE}) with 
$\DG=\Delta$,
and $\rho_F$ is a Gaussian density. We allow a wide class of reaction terms, $F$, so
that the Fisher-KPP equation emerges as a special case.

We remark that such results on (one-step) convergence to reaction diffusion 
equation limits have been
achieved for a variety of interacting particle systems. The vast majority of these
results concern spin systems \comment{REFs}. However,~\cite{flandoli/huang:2021} 
obtain the Fisher-KPP equation (that is equation~(\ref{semilinear PDE}) with 
$\DG=\Delta$ and $F(\varphi)=1-\varphi$) as a scaling limit of interacting diffusions in
$\IR^d$. In that setting a major challenge (also apparent in our models), not
faced in spin systems, is 
the control of concentration of particles. We also note that the scaling 
considered in~\cite{flandoli/huang:2021} is different from that considered
below. In their language we consider only `moderate interactions', meaning that
the number of individuals in the neighbourhood over which we measure local
population density tends to infinity, whereas they also consider the situation in which
that number remains finite; furthermore, whereas we shall consider a scaling in which 
the reproduction mechanism is near-critical
and look at large timescales,~\cite{flandoli/huang:2021}
only scale the number of
particles in the system and the size of the neighbourhood over which indivduals
interact.

The history of a natural population is often only accessible indirectly, 
through patterns of genetic diversity that have been laid down; from 
genetic data, one can try to infer the genealogical trees that relate 
individuals in a sample from the population, and these have been shaped 
by its history. It is therefore of interest
to establish information about the distribution of genealogical trees 
under our population model.
To achieve this we exploit a lookdown construction. 
Lookdown constructions were first introduced in~\cite{donnelly/kurtz:1996} to provide 
a mechanism for retaining information about genealogical relationships between individuals
sampled from a population evolving according to the Moran model when passing to the 
infinite population limit.
Since then, they have been extended to a huge range of models. Of particular
relevance to our work here are the papers of~\cite{kurtz/rodrigues:2011} 
and~\cite{etheridge/kurtz:2019}, in which
lookdown constructions are provided for a wide variety of population models, including
spatially structured branching processes.

@@

Most work has focused on models that caricature the classical Fisher-KPP equation with a 
stochastic term, i.e.
$$dw=\big(\Delta w +sw(1-w)\big)dt +\sqrt{\frac{\alpha(w)}{N}}W(dt,dx),$$
where $W$ is space-time white noise, and $N$ is a measure of the 
local population density. The coefficient
$\alpha(w)$ is generally taken to be either
$w$, corresponding to a superprocess limit, or $w(1-w)$ giving a 
spatial analogue of a Wright-Fisher diffusion. 
For very large $N$, and on suitable timescales, 
starting with the pioneering work of~\cite{brunet/derrida/mueller/munier:2006},
a considerable body of evidence has been amassed to underpin the
conjecture that for this, and a wide class of 
related models, 
genealogies converge to a 
Bolthausen-Sznitman coalescent. %, e.g.~\cite{brunet/derrida/mueller/munier:2007}.
This reflects the fact that, for this equation, 
ancestral lineages become trapped in 
the wavefront, where the growth rate of the population is highest. 
Once there, they will experience 
rapid periods of coalescence 
corresponding to significant proportions of individuals in the front being 
descended from particularly reproductively successful ancestors.

If one replaces the logistic growth term of the classical
Fisher-KPP equation with a nonlinearity that reflects cooperative
behaviour in the population, such as
\begin{equation}
	\label{Birzu nonlinearity}
	wF(w)=w(1-w)(Cw-1),
\end{equation}
then, for sufficiently large $C$ (strong cooperation),
the nature of the deterministic
wave changes from ``pulled'' to ``pushed'', 
\cite{birzu/hallatschek/korolev:2017, birzu/hallatschek/korolev:2021}.
In that setting, the genealogies will be quite different
from the Fisher-KPP case. 
%\comment{Could refer to more recent Birzu et al paper here.}
For example, \cite{etheridge/penington:2022}
show that for a discrete space model corresponding to this 
nonlinearity with $C>2$, after suitable scaling, the genealogy of a
sample converges not to a Bolthausen-Sznitman coalescent, but to
a Kingman coalescent. 
The reason, roughly, is that ancestral lineages
settle to a stationary distribution relative to the position of the 
wavefront which puts very little weight close to the `tip' of the wave, so
that when ancestral lineages meet
it is typically at a location in which population density is
high, where no single ancestor produces a disproportionately large number of 
descendants in a short space of time.

The shape of the wave is not determined solely by the reaction term.
For example, as a result of the nonlinear diffusion, 
for suitable initial conditions, the solution to the one-dimensional 
porous medium equation with logistic growth~(\ref{PME1}) converges to a travelling
wave with a sharp cut-off; i.e., in contrast to the classical 
Fisher KPP equation, the solution at time $t$
vanishes beyond $x=x_0+ct$ for some constant wavespeed
$c>0$, \cite{kamin/rosenau:2004}.
As a first step towards understanding what we should expect in models with
nonlinear diffusion, one can ask about the position of an ancestral lineage
relative to the wavefront in the deterministic models. In 
Section~\ref{ancestral lineages for travelling waves}
we shall see that in our framework, even with logistic growth, the nonlinear diffusion 
corresponding to the porous medium equation results
in a stationary distribution that is concentrated behind the wavefront, leading 
us to conjecture that in the stochastic equation
the cooperative behaviour captured by the nonlinear diffusion will also 
result in a qualitatively different pattern of coalescence to that seen under the 
stochastic Fisher-KPP equation. Indeed, we believe that it should be feasible to 
show that in an appropriate limit one recovers a Kingman coalescent.

@@

Our model is one of individuals distributed across a continuous space 
which we shall take to be $\IR^d$. 
For applications, $d=1$ or $d=2$, but our main results apply more generally.
At time zero, the population is distributed over a bounded region, with 
$\bigO(N)$ individuals per unit area in that region,
so the total number of individuals will also be $\bigO(N)$.
The population changes in continuous time, and
we encode the state of the population at time $t$ by a counting measure $X(t)$,
which assigns one unit of mass to the location of each individual.

@@

We shall write $\mu_\theta(x, X/N)$ for the per-capita death rate of mature individuals
in the population.
In order for the 
population density to change over timescales of order $\theta$, we should like the net 
per capita reproductive rate 
to scale as $1/\theta$. 
In classical models, in which $r$, $\gamma$, and $\mu$ are constant, 
this quantity is simply $r\gamma -\mu$. 
Here, because production of juveniles and their establishment are
mediated by population density measured relative to different points, 
the net reproductive rate will take a more complicated form. 
In particular, the total rate of production of mature offspring by an individual at $x$ will be
\[
	\gamma\big(x,\smooth{\gamma}X(x)/N\big)\int r\big(y, \smooth{r}X(y)/N\big)q_{\theta}(x,dy).
\]
Nonetheless it will be convenient to define the death rate $\mu_\theta$ in terms of 
its deviation from $r\gamma$. 
To this end, we define
the death rate of an individual at $x$ to be
\begin{align} \label{eqn:mu_defn}
    \mu_\theta(x)
    =
    \max\left\{0, 
        r(x, \smooth{r} X(x) / N) \gamma(x, \smooth{\gamma} X(x) / N)
        - \frac{1}{\theta} F(x, \smooth{F} X(x) / N)
    \right\} ,
\end{align}
where $F(x, m) : \IR^d \times \IR_{\ge 0} \to \IR$ and
$\rho_F$ is again a smooth kernel.

@@

\begin{notation}
Expressions like $\gamma(x, \smooth{\gamma} \eta(x))$ will appear repeatedly in what follows.
To make formulae more readable, we overload notation to define
$$
    \gamma(x, \eta) := \gamma(x, \smooth{\gamma} \eta(x)) ,
$$
and similarly write $r(x, \eta)$ for $r(x, \smooth{r} \eta(x))$
and $\mu_\theta(x, \eta)$ for the expression of equation~\eqref{eqn:mu_defn}.
When convenient, we may also suppress the arguments completely,
writing simply $\gamma$, $r$, and $\mu_\theta$ for these quantities.
\end{notation}

\begin{terminology*}
In our prelimiting model,
the population is represented by a point measure in which each
individual is assigned a mass $1/N$. We use the term ``population density'' for
this process, as it is supposed to measure population size relative to a 
nominal occupancy of $N$ individuals per unit area. There is no implication 
that the measure representing the population is absolutely continuous with
respect to Lebesgue measure; indeed in the prelimit it is certainly not.
\end{terminology*}

@@

\begin{definition}[Dispersal generator]
    \label{def:dispersal_generator}
    As above, we define the dispersal kernel,
    $q_\theta(x, dy)$,
    to be the density of a multivariate Gaussian
    with mean $\meanq(x)/\theta$ and covariance matrix $\covq(x)/\theta$
    (although often we omit the dependence of $\meanq$ and $\covq$ on $x$).
    This implies that if we define, for $f : \IR^d \to \IR$,
    \begin{align}
    \DG f(x)
        =
        \sum_{ij} \covq(x)_{ij} \partial_{x_i} \partial_{x_j} f(x)
        + \sum_i \meanq(x)_i \partial_{x_i} f(x)
    \end{align}
    then
    \begin{align}
        \theta \int \left(
            f(y) - f(x)
        \right) q_\theta(x, dy)
    \to \DG f(x) 
        \qquad \text{as } \theta \to \infty .
    \end{align}
    We will also denote the adjoint of $\DG$ by
    \begin{align}
	    \nonumber
    \DG^* f(x)
        &=
        \sum_{ij} \partial_{x_i} \partial_{x_j} (\covq(x)_{ij} f(x))
        - \sum_i \partial_{x_i} (f(x) \meanq(x)_i) 
        \\
	    \label{drift in adjoint}
        &=
        \sum_{ij} \partial_{x_i} \partial_{x_j} (\covq(x)_{ij} f(x))
	    -\nabla f(x)\cdot \meanq(x)-f(x)\nabla\cdot\meanq(x).
        %\text{TODO} .
    \end{align}
\end{definition}

@@

{\bf Dispersal generator:}
We assume that
\begin{enumerate}
    \item $\meanq(x)$, $\covq(x)$ are uniformly bounded in each component;
    \item $\meanq(x)$ and $\covq(x)$ are $\alpha$-H\"older continuous in each component, 
	    for some $\alpha \in (0,1]$;
    \item the operator $\DG$ is uniformly strictly elliptic.
\end{enumerate}

@@

The purpose of the conditions that we have placed on the reproduction parameters is to 
ensure that the net per capita reproduction rate (before time scaling)
is order $1/\theta$. As remarked above, 
because of the non-local reproduction mechanism, it no longer suffices to assume that 
$r(x,\smooth{r}\eta(x))\gamma(x, \smooth{\gamma}\eta(x))-\mu_\theta(x)$ is of order 
$1/\theta$. The conditions that we have imposed are to ensure that the difference 
between the establishment probability of a juvenile born of a parent at $x$ and dispersed
to a position $y$, determined by $q_{\theta}(x,dy)$, and its establishment probability had
it remained at $x$ are also order $1/\theta$.

>>>>

\author{ \begin{small}
\begin{tabular}{ll}                              
Alison M. Etheridge 
 & Thomas G. Kurtz \\   
Department of Statistics & Departments of Mathematics and Statistics\\       
Oxford University & University of Wisconsin - Madison \\                   
24-29 St Giles & 480 Lincoln Drive\\                                                         
Oxford OX1 3LB & Madison, WI  53706-1388\\
UK & USA \\                        
etheridg@stats.ox.ac.uk & kurtz@math.wisc.edu     \\
\url{http://www.stats.ox.ac.uk/~etheridg/} & 
\url{http://www.math.wisc.edu/~kurtz/}  \\       \\
\\
Ian Letter&  Peter L. Ralph 
\\   
Department of Statistics & Departments of Mathematics and Biology\\
Oxford University &University of Oregon\\                   
24-29 St Giles & Fenton Hall\\
Oxford OX1 3LB & Eugene, OR 97403-1222\\
UK & USA \\
restucci@stats.ox.ac.uk  & plr@oregon.edu \\
\url{https://www.stats.ox.ac.uk/~restucci/}&
\url{https://math.uoregon.edu/profile/plr} \\
\\
Terence Tsui 
 &  \\   
Department of Statistics & \\
Oxford University & \\                   
24-29 St Giles& \\
Oxford OX1 3LB & \\
UK & \\
terence.tsui@sjc.ox.ac.uk &      \\
\url{https://www.maths.ox.ac.uk/people/terence.tsui}&  \\
\end{tabular}
\end{small}}

@@

We introduce a broad class of mechanistic spatial models to describe how spatially 
heterogeneous populations live, die, and reproduce.
Individuals are represented by points of a point measure,
whose birth and death rates can depend 
both on spatial position and local population density, defined at a location
to be the convolution of the point measure
with a suitable non-negative integrable kernel
centred on that location. We pass to three different scaling limits:
an interacting superprocess, a nonlocal partial differential equation (PDE), 
and a classical PDE. 
The classical PDE is obtained both by a two-step convergence argument, in which we 
first scale time and population size and
pass to the nonlocal PDE, 
and then scale the kernel that 
determines local population density; and 
in the important special case in which the limit is a 
reaction diffusion equation, directly 
by simultaneously scaling the kernel and the
timescale and population size in our individual based model.

A novelty of our model is that we explicitly
model a juvenile phase: the number of juveniles produced by an individual depends
on local population density at the location of the parent. 
Juvenile offspring are 
thrown off in a (possibly heterogeneous, anisotropic) Gaussian distribution 
around the location of the parent; they then reach (instant) maturity with a probability
that can depend on the local population density
% (possibly measured through convolution with a different kernel from that used in determining fecundity) 
at the location at which they land. 
Although we only record mature individuals, a trace of this two-step description 
remains in our population models, resulting in novel limits in which the spatial
motion is governed by a nonlinear diffusion.

@@

Reproduction of individuals naturally leads to spatial branching process models,
including branching random walk, 
branching Brownian motion, and the Dawson-Watanabe superprocesses.  
However, as a result of the branching assumption (once born, individuals behave 
independently of one another), 
a population evolving according to any of these models will
either die out or grow without bound and, in so doing, can
develop clumps of arbitrarily large density and extent. 
Our starting point here is an individual-based model of a single species 
in continuous space in which
birth, death, and establishment may all depend on local population density
as well as on spatial location,
allowing for stable populations through density-dependent feedback.

Although it is often mathematically convenient to assume that individuals follow
Brownian motion during their lifetime, 
in our model, offspring are thrown off according to some 
spatial distribution centred on the location of the parent
and do not subsequently move.
This is particularly appropriate for modelling plant populations, in which
this dispersal of offspring around the parent is the only source of
spatial motion.

Most models do not distinguish between juveniles and adults, so,
for example, the number of adults produced by a single parent is determined
by the degree of crowding at the location of the parent.
Although we shall similarly only follow the adult population, in formulating the dynamics of the
models we shall distinguish
between production of juveniles, which will depend upon the location of 
the adult, and their successful establishment, which will depend on the
location in which a juvenile lands. The result is that not only the absolute 
number, but also the spatial distribution
around their parent, 
of those offspring that survive to adulthood
will depend upon the local population 
density.

We shall consider three different classes of scaling limits for our model.
The first yields a class of (generalised) superprocesses in which coefficients 
governing both the spatial motion and the branching components of the process can depend
on local population density; the second is a corresponding class of deterministic
non-local differential equations; and the third are classical PDEs.
We measure local population density around a point by convolving with
a smooth kernel $\rho(\cdot)$, which may differ
for the two stages of reproduction. 
When the limiting population process is deterministic,
it is a (weak) solution of an equation of the form
\begin{equation}
	\label{general deterministic limit}
        \partial_t \varphi_t(x)
        =
        r\left(x, \varphi_t \right)
        \DG^* \left[
            \varphi_t(\cdot)
            \gamma\big( x, \varphi_t \big)
        \right](x)
        +
        \varphi_t(x)
        F\left(x, \varphi_t \right)
        ,
\end{equation}
where
$\varphi_t(x)$ can be thought of as the population density at $x$
(although the limit may be a measure without a density), and
$\DG^*$ is (the adjoint of) a strictly uniformly elliptic second order differential operator, typically the Laplacian.
The dependence of each of the terms $r$, $\gamma$, and $F$ on $\varphi$ is only through the local density at $x$,
e.g., $F(x, \varphi) = F(x, \smooth{} \varphi(x))$.
We shall be more specific about the parameters below.

By replacing $\rho$ by
$\rho^\epsilon(\cdot)=\rho(\cdot/\epsilon)/\epsilon^d$, 
we can also scale the ``width'' of the 
region over which we measure local population density. 
When the population follows~(\ref{general deterministic limit}), 
we expect that if we take a second limit of $\epsilon \to 0$,
thus scaling the kernels appearing in $r$, $\gamma$, and $F$ and making interactions are pointwise,
we should recover a nonlinear PDE.
We verify that this is indeed the case in two important examples:
a special case of the porous medium equation with a logistic growth term, in which we 
retain a signal of our two stage approach to modelling in the nonlinear diffusion term
and the limiting equation takes the form 
\begin{equation}
	\label{PME1}
	\partial_t \varphi = \Delta (\varphi^2)+\varphi(1-\varphi);
\end{equation}
and a wide class of semi-linear PDEs of the form 
\begin{equation}
\label{semilinear PDE}
\partial_t \varphi = \DG^*\varphi+ \varphi F(\varphi),
\end{equation}
which includes the Fisher-KPP equation and the Allen-Cahn equation.

It is of interest to understand under what conditions we can replace the two-step limiting
process described above by one in which we simultaneously scale the 
kernels
and the other parameters in 
our population model to arrive at the PDE limit.
This is mathematically much more challenging, but we establish such 
one-step convergence in cases for which the limit is a classical 
reaction diffusion equation of the form~(\ref{semilinear PDE}) with 
$\DG=\Delta$,
and $\rho_F$ is a Gaussian density. We allow a wide class of reaction terms, $F$, so
that the Fisher-KPP equation 
(that is equation~(\ref{semilinear PDE}) with $\DG=\Delta$ and $F(\varphi)=1-\varphi$)
emerges as a special case.

We remark that such results on (one-step) convergence to reaction diffusion 
equation limits have been
achieved for a variety of interacting particle systems. The vast majority of these
results concern spin systems \comment{REFs}. However,~\cite{flandoli/huang:2021} 
obtain the Fisher-KPP equation
as a scaling limit of interacting diffusions in
$\IR^d$. In that setting a major challenge (also apparent in our models), not
faced in spin systems, is 
the control of concentration of particles. We also note that the scaling 
considered in~\cite{flandoli/huang:2021} is different from that considered
below. In their language we consider only `moderate interactions', meaning that
the number of individuals in the neighbourhood over which we measure local
population density tends to infinity, whereas they also consider the situation in which
that number remains finite; furthermore.
On the other hand, we shall consider a scaling in which 
the reproduction mechanism is near-critical and look at large timescales,
while \citet{flandoli/huang:2021} only scale the number of
particles in the system and the size of the neighbourhood over which indivduals interact.

The history of a natural population is often only accessible indirectly, 
through patterns of genetic diversity that have been laid down; from 
genetic data, one can try to infer the genealogical trees that relate 
individuals in a sample from the population, and these have been shaped 
by its history \citep{neigel1993application,kelleher2019inferring}.
It is therefore of interest to establish information about the distribution of genealogical trees 
under our population model.
To achieve this we exploit a lookdown construction. 
Lookdown constructions were first introduced in~\cite{donnelly/kurtz:1996} to provide 
a mechanism for retaining information about genealogical relationships between individuals
sampled from a population evolving according to the Moran model when passing to the 
infinite population limit.
Since then, they have been extended to a huge range of models. Of particular
relevance to our work here are the papers of~\cite{kurtz/rodrigues:2011} 
and~\cite{etheridge/kurtz:2019}, in which
lookdown constructions are provided for a wide variety of population models, including
spatially structured branching processes.

@@

Most work on this topic has focused on models that caricature the classical Fisher-KPP equation with a 
stochastic term, i.e.
$$dw=\big(\Delta w +sw(1-w)\big)dt +\sqrt{\frac{\alpha(w)}{N}}W(dt,dx),$$
where $W$ is space-time white noise, and $N$ is a measure of the 
local population density. The coefficient
$\alpha(w)$ is generally taken to be either
$w$, corresponding to a superprocess limit, or $w(1-w)$ giving a 
spatial analogue of a Wright-Fisher diffusion. 
For very large $N$, and on suitable timescales, 
starting with the pioneering work of~\cite{brunet/derrida/mueller/munier:2006},
a considerable body of evidence has been amassed to underpin the
conjecture that for this, and a wide class of 
related models, 
genealogies converge to a 
Bolthausen-Sznitman coalescent. %, e.g.~\cite{brunet/derrida/mueller/munier:2007}.
This reflects the fact that, for this equation, 
ancestral lineages become trapped in 
the wavefront, where the growth rate of the population is highest. 
Once there, they will experience 
rapid periods of coalescence 
corresponding to significant proportions of individuals in the front being 
descended from particularly reproductively successful ancestors.

If one replaces the logistic growth term of the classical
Fisher-KPP equation with a nonlinearity that reflects cooperative
behaviour in the population, such as
\begin{equation}
	\label{Birzu nonlinearity}
	wF(w)=w(1-w)(Cw-1),
\end{equation}
then, for sufficiently large $C$ (strong cooperation),
the nature of the deterministic
wave changes from ``pulled'' to ``pushed'', 
\citep{birzu/hallatschek/korolev:2017, birzu/hallatschek/korolev:2021}.
In that setting, the genealogies will be quite different
from the Fisher-KPP case. 
%\comment{Could refer to more recent Birzu et al paper here.}
For example, \citet{etheridge/penington:2022}
show that for a discrete space model corresponding to this 
nonlinearity with $C>2$, after suitable scaling, the genealogy of a
sample converges not to a Bolthausen-Sznitman coalescent, but to
a Kingman coalescent. 
The reason, roughly, is that ancestral lineages
settle to a stationary distribution relative to the position of the 
wavefront which puts very little weight close to the `tip' of the wave, so
that when ancestral lineages meet
it is typically at a location in which population density is
high, where no single ancestor produces a disproportionately large number of 
descendants in a short space of time.

The shape of the wave is not determined solely by the reaction term.
For example, as a result of the nonlinear diffusion, 
for suitable initial conditions, the solution to the one-dimensional 
porous medium equation with logistic growth~(\ref{PME1}) converges to a travelling
wave with a sharp cut-off; i.e., in contrast to the classical 
Fisher KPP equation, the solution at time $t$
vanishes beyond $x=x_0+ct$ for some constant wavespeed $c>0$, \citep{kamin/rosenau:2004}.
As a first step towards understanding what we should expect in models with
nonlinear diffusion, one can ask about the position of an ancestral lineage
relative to the wavefront in the deterministic models. In 
Section~\ref{ancestral lineages for travelling waves}
we shall see that in our framework, even with logistic growth, the nonlinear diffusion 
corresponding to the porous medium equation results
in a stationary distribution that is concentrated behind the wavefront, leading 
us to conjecture that in the stochastic equation
the cooperative behaviour captured by the nonlinear diffusion will also 
result in a qualitatively different pattern of coalescence to that seen under the 
stochastic Fisher-KPP equation. Indeed, we believe that it should be feasible to 
show that in an appropriate limit one recovers a Kingman coalescent.

@@

Our model is one of individuals distributed across a continuous space 
which we shall take to be $\IR^d$. 
For applications, $d=1$ or $d=2$
(or even $d=3$ for cells within the body),
but our main results apply more generally.
At time zero, the population is distributed over a bounded region, with 
$\bigO(N)$ individuals per unit area in that region,
so the total number of individuals will also be $\bigO(N)$.
The population changes in continuous time, and
we encode the state of the population at time $t$ by a counting measure $X(t)$,
which assigns one unit of mass to the location of each individual.

@@

We shall write $\mu_\theta(x, X/N)$ for the per-capita death rate of mature individuals
in the population.
In order for the 
population density to change over timescales of order $\theta$, we should like the net 
per capita reproductive rate 
to scale as $1/\theta$. 
In classical models, in which $r$, $\gamma$, and $\mu$ are constant, 
this quantity is simply $r\gamma -\mu$. 
Here, because production of juveniles and their establishment are
mediated by population density measured relative to different points, 
the net reproductive rate will take a more complicated form. 
In particular, the total rate of production of mature offspring by an individual at $x$ will be
\[
	\gamma\big(x,\smooth{\gamma}X(x)/N\big)\int r\big(y, \smooth{r}X(y)/N\big)q_{\theta}(x,dy).
\]
Nonetheless it will be convenient to define the death rate $\mu_\theta$ in terms of 
its deviation from $r\gamma$. 
To this end, we define
the death rate of an individual at $x$ to be
\begin{align} \label{eqn:mu_defn}
    \mu_\theta(x)
    =
    \max\left\{0, 
        r(x, \smooth{r} X(x) / N) \gamma(x, \smooth{\gamma} X(x) / N)
        - \frac{1}{\theta} F(x, \smooth{F} X(x) / N)
    \right\} ,
\end{align}
where this equation defines $F(x, m) : \IR^d \times \IR_{\ge 0} \to \IR$,
the net per capita reproductive rate, scaled by $\theta$,
and $\rho_F$ is again a smooth kernel. 
It will be convenient to work with $F$ instead of $\mu$.

@@

\begin{notation}
Expressions like $\gamma(x, \smooth{\gamma} \eta(x))$ will appear repeatedly in what follows.
To make formulae more readable, we overload notation to define
$$
    \gamma(x, \eta) := \gamma(x, \smooth{\gamma} \eta(x)) ,
$$
and similarly write $r(x, \eta)$ for $r(x, \smooth{r} \eta(x))$,
$F(x, \eta)$ for $F(x, \smooth{F} \eta(x))$,
and $\mu_\theta(x, \eta)$ for the expression of equation~\eqref{eqn:mu_defn}.
When convenient, we may also suppress the arguments completely,
writing simply $\gamma$, $r$, $F$, and $\mu_\theta$ for these quantities.
\end{notation}

\comment{Do we need a "Terminology" heading for this?}
\begin{terminology*}
In our prelimiting model,
the population is represented by a point measure in which each
individual is assigned a mass $1/N$. We use the term ``population density'' for
this process, as it is supposed to measure population size relative to a 
nominal occupancy of $N$ individuals per unit area. There is no implication 
that the measure representing the population is absolutely continuous with
respect to Lebesgue measure; indeed in the prelimit it is certainly not.
\end{terminology*}

@@

\begin{definition}[Dispersal generator]
    \label{def:dispersal_generator}
    As above, we define the dispersal kernel,
    $q_\theta(x, dy)$,
    to be the density of a multivariate Gaussian
    with mean $\meanq(x)/\theta$ and covariance matrix $\covq(x)/\theta$
    (although often we omit the dependence of $\meanq$ and $\covq$ on $x$).
    This implies that if we define, for $f : \IR^d \to \IR$,
    \begin{align}
    \DG f(x)
        =
        \sum_{ij} \covq(x)_{ij} \partial_{x_i} \partial_{x_j} f(x)
        + \sum_i \meanq(x)_i \partial_{x_i} f(x)
    \end{align}
    then
    \begin{align}
        \theta \int \left(
            f(y) - f(x)
        \right) q_\theta(x, dy)
    \to \DG f(x) 
        \qquad \text{as } \theta \to \infty .
    \end{align}
    We will also denote the adjoint of $\DG$ by
    \begin{align}
	    \nonumber
    \DG^* f(x)
        &=
        \sum_{ij} \partial_{x_i} \partial_{x_j} (\covq(x)_{ij} f(x))
        - \sum_i \partial_{x_i} (f(x) \meanq(x)_i) 
        \\
	    \label{drift in adjoint}
        % &=
        % \sum_{ij} \partial_{x_i} \partial_{x_j} (\covq(x)_{ij} f(x))
	    % -\nabla f(x)\cdot \meanq(x)-f(x)\nabla\cdot\meanq(x).
        &=
        \sum_{ij} C_{ij}(x) \partial_{x_i} \partial_{x_j} f(x)
        + \sum_i \left(
            \sum_j \partial_{x_j} C_{ij}(x) - \meanq_i(x)
        \right) \partial_{x_i} f(x)
        \\ & \qquad {}
        + \left(
            \sum_{ij} \partial_{x_i} \partial_{x_j} C_{ij}(x)
            -
            \sum_i \partial_{x_i} \meanq_i(x)
        \right) f(x) .
    \end{align}
\end{definition}

@@

{\bf Dispersal generator:}
We assume that
\begin{enumerate}
    \item $\meanq(x)$ and $\covq(x)$ are
        $\alpha$-H\"older continuous for some $\alpha \in (0, 1]$
        and uniformly bounded
        in each component;
    % \item $\meanq(x)$ and $\covq(x)$ are $\alpha$-H\"older continuous in each component, for some $\alpha \in (0,1]$;
    \item the operator $\DG$ is uniformly strictly elliptic.
\end{enumerate}

@@

We give two more concrete conditions under which the fourth condition is satisfied.
